Relaxed Weak Accelerated Proximal Gradient Method: a Unified Framework for Nesterov's Accelerations

### 2025-10-14 Notebook LaTEX/R-WAPG/presentation-shorter.tex
>w<, changes

% THEMES AND BEAMER SETTINGS ===================================================
% \usetheme{Madrid}
\graphicspath{{.}}
\input{presets/wang/custom_commands.tex} % IMPORT WANG'S LATEX CUSTOM COMMANDS. 
\setbeamertemplate{theorems}[numbered] % ADD NUMBERING TO ALL AMS THEOREMS. 
\setbeamertemplate{footline}[frame number] % ADD PAGE NUMBERS ON BOTTOM. 
\setbeamertemplate{blocks}[rounded][shadow=true]
\setbeamertemplate{navigation symbols}{} 
\setbeamertemplate{itemize subitem}{$\circ$}
% \setbeamercolor{block title}{bg=cyan,fg=black}  % CHANGE THE BLOCK STYLE IN BEAMER. 
% \setbeamercolor{block body}{bg=lime,fg=black} % CHANGE THE BLOCK STYLE IN BEAMER.

@@

\section{Introduction and background}
    \begin{frame}{Nesterov's accelerated gradient in 1983}
        \begin{block}{Nesterov's accelerated gradient in 1983}
            Initialized $x_0 = y_0 \in \RR^n$. 
            The original formulation by Nesterov in 1983 \cite{nesterov_method_1983} has
            {\small
            \begin{align*}
                & x_{k + 1} := y_k - L^{-1}\nabla F(y_k),
                \\
                & t_{k + 1} := 1/2\left(1 + \sqrt{1 + 4t_{k}^2}\right),
                \\
                & \theta_{k + 1} := (t_{k} - 1)/t_{k + 1}, \label{eqn:example-algorithm}
                \\
                & y_{k + 1} := x_{k + 1} + \theta_{k + 1}(x_{k + 1} - x_k).
            \end{align*}    
            }
        \end{block}
        When $F$ is convex and, $L$ Lipschitz smooth. 
        If minimizer $x^+$ exists, then: 
        \begin{itemize}
            \item a convergence rate of $\mathcal O(1/k^2)$ for $(F(x_k) - F(x^+))_{k \ge 1}$;
            \item this convergence rate is optimal in the sense as proposed by Nesterov \cite{nesterov_lectures_2018}.
        \end{itemize}
    \end{frame}
    \begin{frame}{Choices of momentum parameters $t_k, \theta_k$}
        \begin{block}{Here are prominent results in the literature}
            \begin{itemize}
                \item Beck, Teboulle 2009 \cite{beck_fast_2009}, the FISTA paper: $t_k (t_k - 1) \le t_{k - 1}^2$. 
                    \begin{itemize}
                        \item $(F(x_k) - F(x^+))_{k \ge 0}$ converges at rate $\mathcal O(1/k^2)$ when $t_k(t_k - 1) = t_{k - 1}^2$. 
                        \item $\theta_k$ admits no closed form. 
                    \end{itemize}
                \item Chambolle and Dossal 2015 \cite{chambolle_convergence_2015}: $t_k = (k + a -1)/a$ with $a > 2$. 
                    \begin{itemize}
                        \item Retains convergence rate $\mathcal O(1/k^2)$ with $\theta_{k} = \frac{k - 2}{k + a - 1}$. 
                        \item Iterates has weak convergence in Hilbert space. 
                    \end{itemize}
                \item Nesterov \cite{nesterov_lectures_2018} and, Beck \cite{beck_first-order_2017}
                : $\theta_k = \frac{\sqrt{L/\mu} - 1}{\sqrt{L/\mu} + 1}$.
                    \begin{itemize}
                        \item Converges at rate $\mathcal O\left(\left(1 - \sqrt{\mu/L}\right)^k\right)$, if $f$ is also $\mu$ strongly convex with $\mu > 0$. 
                    \end{itemize}
                \item Apidopoulos et al. 2018 \cite{apidopoulos_convergence_2018}: $\theta_k = \frac{n}{n + b}$ for $b \in (0, 3)$. 
                \begin{itemize}
                    \item Has convergence rate $\mathcal O(1/k^{2b/3})$. 
                \end{itemize}
            \end{itemize}    
        \end{block}
    \end{frame}
    \begin{frame}{A major question}
        The following question lead our investigation: 
        \begin{tcolorbox}
            How much can we relax the choice of $\theta_k, t_k$ and still get an upper bound for the convergence rate? 
        \end{tcolorbox}
        We proposed a framework called Relaxed, Weak Accelerated Proximal Gradient (R-WAPG) and, these are what we did: 
        \begin{itemize}
            \item An upper bound of function value with the weakest assumption on momentum parameter.
            \item A unified analysis with strong convexity $\mu \ge 0$. 
            \item Various equivalent representation of R-WAPG are given and, convergence results varified for examples in the literature. 
            \item Numerical experiments are conducted to verify the theory. 
        \end{itemize}
    \end{frame}

\section{The R-WAPG framework}
    \begin{frame}{The assumptions we make throughout}
        \begin{assumption}[Convex smooth plus nonsmooth]
            Let the ambient space be $\RR^n$, equipped with Euclidean inner product and norm. 
            Define $F := f + g$.
            \begin{enumerate}
                \item $f: \RR^n \rightarrow \RR$ is $L$ Lipschitz smooth and $\mu \ge 0$ strongly convex.
                \item $g: \RR^n \rightarrow \overline \RR$ is proper, closed and convex. The extended real is defined as $\overline \RR := \RR \cup \{\infty\}$.
                \item A minimizer exists for the optimization problem: $F^+ = \min_x \left\lbrace f(x) + g(x)\right\rbrace$.
            \end{enumerate}
        \end{assumption}
        \textbf{This assumption is the full scope of the theoretical and practical discussion. }
    \end{frame}
    \begin{frame}{Proximal gradient, gradient mapping}
        \begin{definition}[The proximal gradient operator]
            Define for all $x\in \RR^n$: 
            \begin{align*}
                T_L(y) 
                &:= \argmin_{x \in \RR^n} \left\lbrace
                    g(x) + \langle \nabla f(y), x\rangle + L/2\Vert x - y\Vert^2
                \right\rbrace 
                \\
                &= \left(I + L^{-1}\partial g\right)^{-1}
                \left(I - L^{-1}\nabla f\right)(y),
                \\
                \mathcal G_L(y)
                &:= L(y - T_L(y)).
            \end{align*}
        \end{definition}
        Note, the $I$ here is the identity operator in $\RR^n$. 
        $L > 0$ is the Lipschitz smoothness parameter for $f$. 
    \end{frame}
    \begin{frame}{The R-WAPG sequences}
        We highlight key theoretical results that allowed us to unify convergence rate for various variants of FISTA in the literature.
        \begin{definition}[R-WAPG sequences]\label{def:rwapg-seq}
            Assume $0 \le \mu < L$. 
            Let $(\alpha_k)_{k \ge 0}$ be such that: 
            \begin{align*}
                & \alpha_0 \in (0, 1], \\
                & \alpha_k \in (\mu/L, 1) \quad (\forall k \ge 1). 
            \end{align*}    
            Define $(\rho_k)_{k \ge 0}$
            \begin{align*}
                \rho_k := \frac{\alpha_{k + 1}^2 - (\mu/L) \alpha_{k + 1}}{
                    (1 - \alpha_{k + 1}) \alpha_k^2
                }. 
            \end{align*}
            We call the sequence $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge0}$ the R-WAPG sequences. 
        \end{definition}
    \end{frame}
    \begin{frame}{R-WAPG}
        \begin{definition}[R-WAPG]\label{def:wapg}
            Choose any $x_1 \in \RR^n, v_1 \in \RR^n$.
            Let $(\alpha_k)_{k \ge0}, (\rho_k)_{k \ge 0}$ be given by Definition \ref{def:rwapg-seq}.
            The algorithm generates a sequence of vector $(y_k, x_{k + 1}, v_{k + 1})_{k \ge 1}$ for all $k\ge 1$ by such that:
            \begin{tcolorbox}\vspace{-1em}
                \begin{align*}
                    \gamma_k &\defeq \rho_{k -1}L\alpha_{k - 1}^2,
                    \\
                    L\alpha_k^2 &= (1 - \alpha_k)\gamma_k + \mu \alpha_k, 
                    \\
                    \hat \gamma_{k + 1} & \defeq L\alpha_k^2,
                    \\
                    y_k &=
                    (\gamma_k + \alpha_k \mu)^{-1}(\alpha_k \gamma_k v_k + \hat\gamma_{k + 1} x_k),
                    \\
                    g_k &= \mathcal G_L (y_k),
                    \\
                    v_{k + 1} &=
                    \hat\gamma^{-1}_{k + 1}
                    (\gamma_k(1 - \alpha_k) v_k - \alpha_k g_k + \mu \alpha_k y_k),
                    \\
                    x_{k + 1} &= T_L (y_k).
                \end{align*}
            \end{tcolorbox}
        \end{definition}
    \end{frame}

@@

            If the sequence $(y_k, v_k, x_k)_{k \ge 1}$ is produced by R-WAPG (Definition \ref{def:wapg}), and $\alpha_0 = 1, x_1 = v_1$, then they satisfy for all $k\ge 1$: 

>>>>

% THEMES AND BEAMER SETTINGS ===================================================
% \usetheme{Madrid}
\graphicspath{{.}}
\input{presets/wang/custom_commands.tex} % IMPORT WANG'S LATEX CUSTOM COMMANDS. 
\setbeamertemplate{theorems}[numbered] % ADD NUMBERING TO ALL AMS THEOREMS. 
\setbeamertemplate{footline}[frame number] % ADD PAGE NUMBERS ON BOTTOM. 
\setbeamertemplate{blocks}[rounded][shadow=false]
\setbeamertemplate{navigation symbols}{} 
\setbeamertemplate{itemize subitem}{$\circ$}
% \setbeamercolor{block title}{bg=cyan,fg=black}  % CHANGE THE BLOCK STYLE IN BEAMER. 
% \setbeamercolor{block body}{bg=lime,fg=black} % CHANGE THE BLOCK STYLE IN BEAMER.

@@

\section{Introduction and background}
    \begin{frame}{Nesterov's accelerated gradient in 1983}
        \begin{block}{Nesterov's accelerated gradient in 1983}
            Initialized $x_0 = y_0 \in \RR^n$. 
            The original formulation by Nesterov in 1983 \cite{nesterov_method_1983} has
                & x_{k + 1} := y_k - L^{-1}\nabla F(y_k),
                & t_{k + 1} := 1/2\left(1 + \sqrt{1 + 4t_{k}^2}\right),
                & \theta_{k + 1} := (t_{k} - 1)/t_{k + 1}, \label{eqn:example-algorithm}
                & y_{k + 1} := x_{k + 1} + \theta_{k + 1}(x_{k + 1} - x_k).
        \end{block}
        When $F$ is convex and, $L$ Lipschitz smooth. 
        If minimizer $x^+$ exists, then: 
        \begin{itemize}
            \item a convergence rate of $\mathcal O(1/k^2)$ for $(F(x_k) - F(x^+))_{k \ge 1}$;
            \item this convergence rate is optimal in the sense as proposed by Nesterov \cite{nesterov_lectures_2018}.
        \end{itemize}
    \begin{frame}{Choices of momentum parameters $t_k, \theta_k$}
        \begin{block}{Here are prominent results in the literature}
            \begin{itemize}
                \item Beck, Teboulle 2009 \cite[Theorem 4.4]{beck_fast_2009-1}, the FISTA paper: $t_k (t_k - 1) \le t_{k - 1}^2$. 
                    \begin{itemize}
                        \item $(F(x_k) - F(x^+))_{k \ge 0}$ converges at rate $\mathcal O(1/k^2)$ when $t_k(t_k - 1) = t_{k - 1}^2$. 
                        \item $\theta_k$ admits no closed form. 
                    \end{itemize}
                \item Chambolle and Dossal 2015 \cite[Theorem 4.1]{chambolle_convergence_2015}: $t_k = (k + a -1)/a$ with $a > 2$. 
                    \begin{itemize}
                        \item Retains convergence rate $\mathcal O(1/k^2)$ with $\theta_{k} = \frac{k - 2}{k + a - 1}$. 
                        \item Iterates converge weakly in Hilbert space. 
                    \end{itemize}
                \item Nesterov \cite[Theorem 2.2.3]{nesterov_lectures_2018} and, Beck \cite[Theorem 10.42]{beck_first-order_2017}
                : $\theta_k = \left(\sqrt{L/\mu} - 1\right)\left(\sqrt{L/\mu} + 1\right)^{-1}$.
                    \begin{itemize}
                        \item Converges at rate $\mathcal O\left(\left(1 - \sqrt{\mu/L}\right)^k\right)$, if $f$ is also $\mu$ strongly convex with $\mu > 0$. 
                    \end{itemize}
                \item Apidopoulos et al. 2018 \cite{apidopoulos_convergence_2018}: $\theta_k = \frac{k}{k + b}$ for $b \in (0, 3)$. 
                \begin{itemize}
                    \item Has convergence rate $\mathcal O(1/k^{2b/3})$. 
                \end{itemize}
            \end{itemize}    
        \end{block}
    \begin{frame}{A major question}
        The following question lead our investigation: 
        \begin{tcolorbox}
            How much can we relax the choice of $\theta_k, t_k$ and still get an upper bound for the convergence rate? 
        \end{tcolorbox}
        We proposed a framework called Relaxed, Weak Accelerated Proximal Gradient (R-WAPG) and, these are what we did: 
        \begin{itemize}
            \item An upper bound of function value with the weakest assumption on momentum parameter.
            \item A unified analysis with strong convexity $\mu \ge 0$. 
            \item Various equivalent representation of R-WAPG are given and, convergence results verified for examples in the literature. 
            \item Numerical experiments are conducted to verify the theory. 
        \end{itemize}

\section{The R-WAPG framework}
    \begin{frame}{The assumptions we make throughout}
        \begin{assumption}[Convex smooth plus nonsmooth]
            Let the ambient space be $\RR^n$, equipped with Euclidean inner product and norm. 
            Define $F := f + g$.
            \begin{enumerate}
                \item $f: \RR^n \rightarrow \RR$ is $L$ Lipschitz smooth and $\mu \ge 0$ strongly convex.
                \item $g: \RR^n \rightarrow \overline \RR$ is proper, closed and convex. The extended real is defined as $\overline \RR := \RR \cup \{\infty\}$.
                \item A minimizer $x^+$ exists for the optimization problem: $F^+ = \min_x \left\lbrace f(x) + g(x)\right\rbrace$.
            \end{enumerate}
        \end{assumption}
        \textbf{This assumption is the full scope of the theoretical and practical discussion. }
    \end{frame}
    \begin{frame}{Proximal gradient, gradient mapping}
        \begin{definition}[The proximal gradient operator]
            Define for all $x\in \RR^n$: 
            \begin{align*}
                T_L(y) 
                &:= \argmin_{x \in \RR^n} \left\lbrace
                    g(x) + \langle \nabla f(y), x\rangle + L/2\Vert x - y\Vert^2
                \right\rbrace 
                \\
                &= \left(I + L^{-1}\partial g\right)^{-1}
                \left(I - L^{-1}\nabla f\right)(y),
                \\
                \mathcal G_L(y)
                &:= L(y - T_L(y)).
            \end{align*}
        \end{definition}
        Note, the $I$ here is the identity operator in $\RR^n$. 
        $L > 0$ is the Lipschitz smoothness parameter for $f$. 
    \end{frame}
    \begin{frame}{The R-WAPG sequences}
        \begin{definition}[R-WAPG sequences]\label{def:rwapg-seq}
            Assume $0 \le \mu < L$. 
            Let $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge 0}$ be such that: 
            \begin{enumerate}
                \item $\alpha_0 \in (0, 1]$, 
                \item $\alpha_k \in (\mu/L, 1) \quad (\forall k \ge 1)$, 
                \item $\rho_k := \frac{\alpha_{k + 1}^2 - (\mu/L) \alpha_{k + 1}}{(1 - \alpha_{k + 1}) \alpha_k^2}$ for all $k \ge 0$. 
            \end{enumerate}  
            We call the sequences $(\alpha_k)_{k \ge 0}, (\rho_k)_{k \ge0}$ the R-WAPG sequences. 
        \end{definition}
    \end{frame}
    \begin{frame}{R-WAPG}
        \begin{definition}[R-WAPG]\label{def:wapg}
            Choose any $x_1 \in \RR^n, v_1 \in \RR^n$.
            Let $(\alpha_k)_{k \ge0}, (\rho_k)_{k \ge 0}$ be given by Definition \ref{def:rwapg-seq}.
            The algorithm generates a sequence of vector $(y_k, x_{k + 1}, v_{k + 1})_{k \ge 1}$ for all $k\ge 1$ by such that:
            \begin{tcolorbox}\vspace{-1em}
                \begin{align*}
                    \gamma_k &\defeq \rho_{k -1}L\alpha_{k - 1}^2,
                    \\
                    L\alpha_k^2 &= (1 - \alpha_k)\gamma_k + \mu \alpha_k, 
                    \\
                    \hat \gamma_{k + 1} & \defeq L\alpha_k^2,
                    \\
                    y_k &=
                    (\gamma_k + \alpha_k \mu)^{-1}(\alpha_k \gamma_k v_k + \hat\gamma_{k + 1} x_k),
                    \\
                    v_{k + 1} &=
                    \hat\gamma^{-1}_{k + 1}
                    \left(\gamma_k(1 - \alpha_k) v_k - \alpha_k \mathcal G_L (y_k) + \mu \alpha_k y_k\right),
                    \\
                    x_{k + 1} &= T_L (y_k).
                \end{align*}
            \end{tcolorbox}
        \end{definition}
    \end{frame}

@@

            If the sequence $(y_k, v_k, x_k)_{k \ge 1}$ is produced by R-WAPG (Definition \ref{def:wapg}), and $\alpha_0 = 1, x_1 = v_1$. 
            Then they satisfy for all $k\ge 1$: 
